The forestecology R package for fitting and assessing neighborhood models of the effect of interspecific competition on the growth of trees

### 2021-08-09 paper/paper.Rmd
set eval = TRUE + ran styler

```{r scbi-trees, out.width="66%", fig.cap="Step 1 - Compute growth of trees based on census data. A map of the growth of a random sample of 500 trees from a 9 ha subsection of the Smithsonian Conservation Biology Institute (SCBI) forest plot."}
ggplot() +
  geom_sf(data = growth_scbi %>% sample_n(500), aes(size = growth)) + 
  scale_size_binned(limits = c(0.1, 1)) +
  labs(size = expression(paste(Growth, " (cm ",y^{-1},')')) )
```

@@

```{r internal, eval = FALSE, echo = FALSE}
# Internal: after removing trees in buffer and have NA for growth, there is one
# tree that is not included in focal_vs_comp_scbi b/c it has no neighbors within
# comp_dist = 7.5m of it
growth_scbi %>% 
  filter(!buffer & !is.na(growth)) %>% 
  filter(!stemID %in% focal_vs_comp_scbi$focal_ID)

@@

```{r appendix1-code, eval = FALSE}
library(tidyverse)
library(lubridate)
library(here)
library(sf)
library(viridis)
library(forestecology)
library(blockCV)
library(tictoc)

# Compute growth of trees based on census data -------------------------
census_2013_scbi <- here("paper/scbi.stem2.csv") %>%
  read_csv() %>%
  select(stemID, sp, date = ExactDate, gx, gy, dbh, codes, status) %>%
  mutate(
    date = mdy(date),
    dbh = as.numeric(dbh)/10
  ) %>%
  filter(gx < 300, between(gy, 300, 600))

census_2018_scbi <- here("paper/scbi.stem3.csv") %>%
  read_csv() %>%
  select(stemID, sp, date = ExactDate, gx, gy, dbh, codes, status) %>%
  mutate(
    date = mdy(date),
    dbh = as.numeric(dbh)/10
  ) %>%
  filter(gx < 300, between(gy, 300, 600))

@@

# Make predictions and compute RMSE
observed_RMSE <- focal_vs_comp_scbi %>%
  mutate(growth_hat = 
           predict(comp_bayes_lm_scbi, focal_vs_comp_scbi)) %>%
  rmse(truth = growth, estimate = growth_hat) %>%
  pull(.estimate)

@@

# 3. Permutation distribution: RMSE with no cross-validation -----------
# Compute num_shuffle permutation test statistics
shuffle_RMSE <- numeric(length = num_shuffle)
for(j in 1:num_shuffle){
  # Fit model (compute posterior parameters) with shuffling
  comp_bayes_lm_scbi <- focal_vs_comp_scbi %>%
    comp_bayes_lm(prior_param = NULL, run_shuffle = TRUE)

@@

# Compute num_shuffle permutation test statistics
for(j in 1:num_shuffle){
  # Compute and save RMSE
  shuffle_RMSE_CV[j] <- focal_vs_comp_scbi %>%
    run_cv(comp_dist = comp_dist, blocks = blocks_scbi, 
           run_shuffle = TRUE) %>%
    rmse(truth = growth, estimate = growth_hat) %>%
    pull(.estimate)

@@

# Visualize results ----------------------------------------------------
model_comp <- bind_rows(
  model_comp_tbl %>% 
    select(run_time, observed = observed_RMSE, 
           shuffle = shuffle_RMSE) %>% 
    mutate(CV = FALSE),
  model_comp_tbl %>% 
    select(run_time, observed = observed_RMSE_CV, 
           shuffle = shuffle_RMSE_CV) %>% 
    mutate(CV = TRUE)
) %>%
  gather(type, RMSE, -c(run_time, CV))

@@

cv_plot <- ggplot() +
  geom_vline(data = model_comp_observed, 
             aes(xintercept = RMSE, col = CV), 
             linetype = "dashed", show.legend = F) +
  geom_histogram(data = model_comp_shuffle, 
                 aes(x = RMSE, fill = CV), bins = 50) +
  labs(
    fill = "Cross-validated?",
    x = expression(paste("RMSE (cm ", y^{-1}, ")"))
  ) +
  scale_color_viridis(discrete = TRUE, option = "D") +
  scale_fill_viridis(discrete = TRUE) +
  theme_light()
cv_plot

filename %>%
  str_c(".pdf") %>%
  ggsave(plot = cv_plot, width = 16/2, height = 9/2)
```

@@

```{r appendix2-code, eval = FALSE}
census_2013_scbi <- read_csv("scbi.stem2.csv") %>%
  select(stemID, sp, date = ExactDate, gx, gy, dbh, codes, status) %>%
  mutate(
    date = mdy(date),
    dbh = as.numeric(dbh)/10
  ) %>%
  filter(gx < 400, gy < 400)

census_2018_scbi <- read_csv("scbi.stem3.csv") %>%
  select(stemID, sp, date = ExactDate, gx, gy, dbh, codes, status) %>%
  mutate(
    date = mdy(date),
    dbh = as.numeric(dbh)/10
  ) %>%
  filter(gx < 400, gy < 400)

growth_scbi <-
  compute_growth(
    census_1 = census_2013_scbi,
    census_2 = census_2018_scbi %>% filter(!str_detect(codes, "R")),
    id = "stemID"
  ) %>%
  # make all species the same, needs to be factor with at least two levels
  mutate(sp = factor('A', levels = c('A', 'B')) ) %>% 
  # Compute basal area:
  mutate(basal_area = 0.0001 * pi * (dbh1 / 2)^2)

@@

# Associate each observation to a fold
spatial_block_scbi <- 
  spatialBlock(speciesData = growth_scbi, k = n_fold, 
               selection = "systematic", blocks = blocks_scbi, 
               showBlocks = FALSE, verbose = FALSE)

@@

mult_dist_comp <- tibble(
  dist = c(5 ,6.25,  7.5, 8.75,  10),
  rmse = 0
)

for (i in 1:length(mult_dist_comp$dist)){
  comp_dist <- mult_dist_comp$dist[i]

@@

mult_dist_comp %>%
  ggplot(aes(dist,rmse)) +
  geom_point() +
  xlab('Distance (m)') +
  ylab('RMSE (cm)')
```

@@

```{r compar-pred-var}
census_2013_scbi <- read_csv("scbi.stem2.csv") %>%
  select(stemID, sp, date = ExactDate, gx, gy, dbh, codes, status) %>%
  mutate(
    date = mdy(date),
    dbh = as.numeric(dbh)/10
  ) %>%
  filter(gx < 300, between(gy, 300, 600))

census_2018_scbi <- read_csv("scbi.stem3.csv") %>%
  select(stemID, sp, date = ExactDate, gx, gy, dbh, codes, status) %>%
  mutate(
    date = mdy(date),
    dbh = as.numeric(dbh)/10
  ) %>%
  filter(gx < 300, between(gy, 300, 600))

growth_scbi <-
  compute_growth(
    census_1 = census_2013_scbi,
    census_2 = census_2018_scbi %>% filter(!str_detect(codes, "R")),
    id = "stemID"
  ) %>%
  left_join(sp_info, by = "sp") %>% 
  mutate(sp = as.factor(sp),
        basal_area = 0.0001 * pi * (dbh1 / 2)^2,
         agb = get_biomass(
                  dbh = dbh1,
                  genus = genus,
                  species = species,
                  coords = c(-78.2, 38.9)
    ))

@@

focal_vs_comp_ba <- growth_scbi %>%
    create_focal_vs_comp(comp_dist = comp_dist, 
                         blocks = blocks_scbi, 
                         id = "stemID", 
                         comp_x_var = "basal_area") %>%
    run_cv(comp_dist = comp_dist, blocks = blocks_scbi)

focal_vs_comp_agb <- growth_scbi %>%
    create_focal_vs_comp(comp_dist = comp_dist, 
                         blocks = blocks_scbi, 
                         id = "stemID", 
                         comp_x_var = "agb") %>%
    run_cv(comp_dist = comp_dist, blocks = blocks_scbi)

@@

  focal_vs_comp_agb %>%
    rmse(truth = growth, estimate = growth_hat) %>%
    pull(.estimate)

```

@@

```{r compare-group-var}

census_2013_scbi <- read_csv("scbi.stem2.csv") %>%
  select(stemID, sp, date = ExactDate, gx, gy, dbh, codes, status) %>%
  mutate(
    date = mdy(date),
    dbh = as.numeric(dbh)/10
  ) %>%
  filter(gx < 300, between(gy, 300, 600))

census_2018_scbi <- read_csv("scbi.stem3.csv") %>%
  select(stemID, sp, date = ExactDate, gx, gy, dbh, codes, status) %>%
  mutate(
    date = mdy(date),
    dbh = as.numeric(dbh)/10
  ) %>%
  filter(gx < 300, between(gy, 300, 600))

growth_scbi_sp <-
  compute_growth(
    census_1 = census_2013_scbi,
    census_2 = census_2018_scbi %>% filter(!str_detect(codes, "R")),
    id = "stemID"
  ) %>%
  mutate(sp = as.factor(sp),
        basal_area = 0.0001 * pi * (dbh1 / 2)^2 )

growth_scbi_can_pos <-
  compute_growth(
    census_1 = census_2013_scbi,
    census_2 = census_2018_scbi %>% filter(!str_detect(codes, "R")),
    id = "stemID"
  ) %>%
  left_join(sp_info, by = "sp") %>% 
  mutate(
    canopy_position = str_replace(canopy_position, ' ', '_'),
    canopy_position = str_replace(canopy_position, ',', ''),
    canopy_position = ifelse(is.na(canopy_position), 'shrub_layer', canopy_position),
    sp = as.factor(canopy_position),
    basal_area = 0.0001 * pi * (dbh1 / 2)^2 )

@@

focal_vs_comp_sp <- growth_scbi_sp %>%
    create_focal_vs_comp(comp_dist = comp_dist, 
                         blocks = blocks_scbi, 
                         id = "stemID", 
                         comp_x_var = "basal_area") %>%
    run_cv(comp_dist = comp_dist, blocks = blocks_scbi)

focal_vs_comp_can_pos <- growth_scbi_can_pos %>%
    create_focal_vs_comp(comp_dist = comp_dist, 
                         blocks = blocks_scbi, 
                         id = "stemID", 
                         comp_x_var = "basal_area") %>%
    run_cv(comp_dist = comp_dist, blocks = blocks_scbi)

@@

```{r can-pos-comp}

  fit_mod_can_pos <- growth_scbi_can_pos %>%
    create_focal_vs_comp(comp_dist = comp_dist, 
                         blocks = blocks_scbi, 
                         id = "stemID", 
                         comp_x_var = "basal_area")  %>%
    comp_bayes_lm(prior_param = NULL)

  autoplot(fit_mod_can_pos, type = "competition")

```

>>>>

```{r scbi-trees, out.width="66%", fig.cap="Step 1 - Compute growth of trees based on census data. A map of the growth of a random sample of 500 trees from a 9 ha subsection of the Smithsonian Conservation Biology Institute (SCBI) forest plot."}
ggplot() +
  geom_sf(data = growth_scbi %>% sample_n(500), aes(size = growth)) +
  scale_size_binned(limits = c(0.1, 1)) +
  labs(size = expression(paste(Growth, " (cm ", y^{-1}, ")")))
```

@@

```{r internal, echo = FALSE}
# Internal: after removing trees in buffer and have NA for growth, there is one
# tree that is not included in focal_vs_comp_scbi b/c it has no neighbors within
# comp_dist = 7.5m of it
growth_scbi %>% 
  filter(!buffer & !is.na(growth)) %>% 
  filter(!stemID %in% focal_vs_comp_scbi$focal_ID)

@@

```{r appendix1-code, eval=FALSE}
library(tidyverse)
library(lubridate)
library(here)
library(sf)
library(viridis)
library(forestecology)
library(blockCV)
library(tictoc)

# Compute growth of trees based on census data -------------------------
census_2013_scbi <- here("paper/scbi.stem2.csv") %>%
  read_csv() %>%
  select(stemID, sp, date = ExactDate, gx, gy, dbh, codes, status) %>%
  mutate(
    date = mdy(date),
    dbh = as.numeric(dbh) / 10
  ) %>%
  filter(gx < 300, between(gy, 300, 600))

census_2018_scbi <- here("paper/scbi.stem3.csv") %>%
  read_csv() %>%
  select(stemID, sp, date = ExactDate, gx, gy, dbh, codes, status) %>%
  mutate(
    date = mdy(date),
    dbh = as.numeric(dbh) / 10
  ) %>%
  filter(gx < 300, between(gy, 300, 600))

@@

# Make predictions and compute RMSE
observed_RMSE <- focal_vs_comp_scbi %>%
  mutate(
    growth_hat =
      predict(comp_bayes_lm_scbi, focal_vs_comp_scbi)
  ) %>%
  rmse(truth = growth, estimate = growth_hat) %>%
  pull(.estimate)

@@

# 3. Permutation distribution: RMSE with no cross-validation -----------
# Compute num_shuffle permutation test statistics
shuffle_RMSE <- numeric(length = num_shuffle)
for (j in 1:num_shuffle) {
  # Fit model (compute posterior parameters) with shuffling
  comp_bayes_lm_scbi <- focal_vs_comp_scbi %>%
    comp_bayes_lm(prior_param = NULL, run_shuffle = TRUE)

@@

# Compute num_shuffle permutation test statistics
for (j in 1:num_shuffle) {
  # Compute and save RMSE
  shuffle_RMSE_CV[j] <- focal_vs_comp_scbi %>%
    run_cv(
      comp_dist = comp_dist, blocks = blocks_scbi,
      run_shuffle = TRUE
    ) %>%
    rmse(truth = growth, estimate = growth_hat) %>%
    pull(.estimate)

@@

# Visualize results ----------------------------------------------------
model_comp <- bind_rows(
  model_comp_tbl %>%
    select(run_time,
      observed = observed_RMSE,
      shuffle = shuffle_RMSE
    ) %>%
    mutate(CV = FALSE),
  model_comp_tbl %>%
    select(run_time,
      observed = observed_RMSE_CV,
      shuffle = shuffle_RMSE_CV
    ) %>%
    mutate(CV = TRUE)
) %>%
  gather(type, RMSE, -c(run_time, CV))

@@

cv_plot <- ggplot() +
  geom_vline(
    data = model_comp_observed,
    aes(xintercept = RMSE, col = CV),
    linetype = "dashed", show.legend = F
  ) +
  geom_histogram(
    data = model_comp_shuffle,
    aes(x = RMSE, fill = CV), bins = 50
  ) +
  labs(
    fill = "Cross-validated?",
    x = expression(paste("RMSE (cm ", y^{
      -1
    }, ")"))
  ) +
  scale_color_viridis(discrete = TRUE, option = "D") +
  scale_fill_viridis(discrete = TRUE) +
  theme_light()
cv_plot

filename %>%
  str_c(".pdf") %>%
  ggsave(plot = cv_plot, width = 16 / 2, height = 9 / 2)
```

@@

```{r appendix2-code}
census_2013_scbi <- read_csv("scbi.stem2.csv") %>%
  select(stemID, sp, date = ExactDate, gx, gy, dbh, codes, status) %>%
  mutate(
    date = mdy(date),
    dbh = as.numeric(dbh) / 10
  ) %>%
  filter(gx < 400, gy < 400)

census_2018_scbi <- read_csv("scbi.stem3.csv") %>%
  select(stemID, sp, date = ExactDate, gx, gy, dbh, codes, status) %>%
  mutate(
    date = mdy(date),
    dbh = as.numeric(dbh) / 10
  ) %>%
  filter(gx < 400, gy < 400)

growth_scbi <-
  compute_growth(
    census_1 = census_2013_scbi,
    census_2 = census_2018_scbi %>% filter(!str_detect(codes, "R")),
    id = "stemID"
  ) %>%
  # make all species the same, needs to be factor with at least two levels
  mutate(sp = factor("A", levels = c("A", "B"))) %>%
  # Compute basal area:
  mutate(basal_area = 0.0001 * pi * (dbh1 / 2)^2)

@@

@@

mult_dist_comp <- tibble(
  dist = c(5, 6.25, 7.5, 8.75, 10),
  rmse = 0
)

for (i in 1:length(mult_dist_comp$dist)) {
  comp_dist <- mult_dist_comp$dist[i]

@@

mult_dist_comp %>%
  ggplot(aes(dist, rmse)) +
  geom_point() +
  xlab("Distance (m)") +
  ylab("RMSE (cm)")
```

@@

```{r compar-pred-var}
census_2013_scbi <- read_csv("scbi.stem2.csv") %>%
  select(stemID, sp, date = ExactDate, gx, gy, dbh, codes, status) %>%
  mutate(
    date = mdy(date),
    dbh = as.numeric(dbh) / 10
  ) %>%
  filter(gx < 300, between(gy, 300, 600))

census_2018_scbi <- read_csv("scbi.stem3.csv") %>%
  select(stemID, sp, date = ExactDate, gx, gy, dbh, codes, status) %>%
  mutate(
    date = mdy(date),
    dbh = as.numeric(dbh) / 10
  ) %>%
  filter(gx < 300, between(gy, 300, 600))

growth_scbi <-
  compute_growth(
    census_1 = census_2013_scbi,
    census_2 = census_2018_scbi %>% filter(!str_detect(codes, "R")),
    id = "stemID"
  ) %>%
  left_join(sp_info, by = "sp") %>%
  mutate(
    sp = as.factor(sp),
    basal_area = 0.0001 * pi * (dbh1 / 2)^2,
    agb = get_biomass(
      dbh = dbh1,
      genus = genus,
      species = species,
      coords = c(-78.2, 38.9)
    )
  )

@@

focal_vs_comp_ba <- growth_scbi %>%
  create_focal_vs_comp(
    comp_dist = comp_dist,
    blocks = blocks_scbi,
    id = "stemID",
    comp_x_var = "basal_area"
  ) %>%
  run_cv(comp_dist = comp_dist, blocks = blocks_scbi)

focal_vs_comp_agb <- growth_scbi %>%
  create_focal_vs_comp(
    comp_dist = comp_dist,
    blocks = blocks_scbi,
    id = "stemID",
    comp_x_var = "agb"
  ) %>%
  run_cv(comp_dist = comp_dist, blocks = blocks_scbi)

@@

focal_vs_comp_agb %>%
  rmse(truth = growth, estimate = growth_hat) %>%
  pull(.estimate)
```

@@

```{r compare-group-var}
census_2013_scbi <- read_csv("scbi.stem2.csv") %>%
  select(stemID, sp, date = ExactDate, gx, gy, dbh, codes, status) %>%
  mutate(
    date = mdy(date),
    dbh = as.numeric(dbh) / 10
  ) %>%
  filter(gx < 300, between(gy, 300, 600))

census_2018_scbi <- read_csv("scbi.stem3.csv") %>%
  select(stemID, sp, date = ExactDate, gx, gy, dbh, codes, status) %>%
  mutate(
    date = mdy(date),
    dbh = as.numeric(dbh) / 10
  ) %>%
  filter(gx < 300, between(gy, 300, 600))

growth_scbi_sp <-
  compute_growth(
    census_1 = census_2013_scbi,
    census_2 = census_2018_scbi %>% filter(!str_detect(codes, "R")),
    id = "stemID"
  ) %>%
  mutate(
    sp = as.factor(sp),
    basal_area = 0.0001 * pi * (dbh1 / 2)^2
  )

growth_scbi_can_pos <-
  compute_growth(
    census_1 = census_2013_scbi,
    census_2 = census_2018_scbi %>% filter(!str_detect(codes, "R")),
    id = "stemID"
  ) %>%
  left_join(sp_info, by = "sp") %>%
  mutate(
    canopy_position = str_replace(canopy_position, " ", "_"),
    canopy_position = str_replace(canopy_position, ",", ""),
    canopy_position = ifelse(is.na(canopy_position), "shrub_layer", canopy_position),
    sp = as.factor(canopy_position),
    basal_area = 0.0001 * pi * (dbh1 / 2)^2
  )

@@

focal_vs_comp_sp <- growth_scbi_sp %>%
  create_focal_vs_comp(
    comp_dist = comp_dist,
    blocks = blocks_scbi,
    id = "stemID",
    comp_x_var = "basal_area"
  ) %>%
  run_cv(comp_dist = comp_dist, blocks = blocks_scbi)

focal_vs_comp_can_pos <- growth_scbi_can_pos %>%
  create_focal_vs_comp(
    comp_dist = comp_dist,
    blocks = blocks_scbi,
    id = "stemID",
    comp_x_var = "basal_area"
  ) %>%
  run_cv(comp_dist = comp_dist, blocks = blocks_scbi)

@@

```{r can-pos-comp}
fit_mod_can_pos <- growth_scbi_can_pos %>%
  create_focal_vs_comp(
    comp_dist = comp_dist,
    blocks = blocks_scbi,
    id = "stemID",
    comp_x_var = "basal_area"
  ) %>%
  comp_bayes_lm(prior_param = NULL)
autoplot(fit_mod_can_pos, type = "competition")
```
